---
title: "ProbabilityIntegralTransform"
output: html_document
date: "2026-04-15"
---

# Probability Integral Transform:

This is just to help me understand the basic theory on why the p-values should appear uniform in the continuous case and contrast this to the discrete case. The proof is straightforward [here](https://en.wikipedia.org/wiki/Probability_integral_transform){.uri}. In this file, I show how the p-values appear when the null distribution is correctly and incorrectly identified, and what happens to the structure depending on how incorrectly it has been specified. In the final code chunk, I also show why the discrete distribution has the property that $p_0^- \ne 1-p_0^+$ since the outcomes have probability mass, which doesn't occur in the continuous case.

```{r}
# The simple continuous case.

unifs <- runif(10000000)
zscore <- qnorm(unifs)

hist(unifs)
hist(zscore)
```

```{r}
# Assume the true distribution is N(1,2), but we assumed it to be N(0,2)

# 1) Simulate values from the N(1,2) but match them back to a N(0,2) CDF
n <- 10000
actual <- rnorm(n, mean = 1, sd = sqrt(2))
hist(actual)

# 2) Match back to incorrect CDF: N(0, 2)
incorrect.CDF <- pnorm(actual, mean = 0, sd = sqrt(2), lower.tail = FALSE)

# 3) View the CDF distribution
hist(incorrect.CDF)

# 4) Compare this to if the initial assumption was correct
correct.CDF <- pnorm(actual, mean = 1, sd = sqrt(2), lower.tail = FALSE)
hist(correct.CDF)

# Summary:
# i) If the null distribution is correctly identified, the p-values should be Uniform
# ii) You must match the data-generating-process to the correct CDF
# iii) If we underestimate the mean, the p-values will be right-skewed
# iv) Increasing the incorrect standard deviation reduces the skewness
```

```{r}
# For a discrete RV, the left tail is not always equal to 1 - right.tail

library(ggplot2)

discrete <- 
  data.frame(x = as.factor(c(1,2,3,4,5)),
             pmass = c(0.1,0.2,0.2,0.1,0.4))

ggplot(data = discrete)+
       geom_bar(aes(x=x, y=pmass), stat = "identity", width = 0.5)

discrete$left.tail <- cumsum(discrete$pmass)
discrete$right.tail <- rev(cumsum(rev(discrete$pmass)))
discrete$one.minus.right <- 1 - discrete$right.tail
discrete$check <- discrete$left.tail == discrete$one.minus.right

discrete
```
